{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-19 thesis/intro.tex
slight revise concl
daves final revisions to intro

My favourite way to explain concurrency to a non-technical audience
%(e.g., family members)
is by analogy with unlocking car doors.
In many %makes and
models of cars, the door handle and locking mechanism are intertwined
in such a way that the door cannot unlock while its handle is being pulled~\cite{car-doors}.
Most car users, then,
%are familiar with
have suffered the experience of
trying to open the passenger door
at exactly the same moment as the driver turns the key in the lock:
%not only does the door not open, but
it fails to unlock, and
they must beseech the driver to turn the key again!
If they had pulled the handle any later, it would have opened as normal;
any earlier, and it would have been as though not having the key at all,
but would at least have unlocked properly at the first turn of the key later.
This violation of expected behaviour -- that the key need not be turned twice --
is a {\em bug},
and its dependence on the two travelers' actions {\em interleaving}
%with the driver's actions
in a particular way
makes it a {\em concurrency bug}.
This example involves only three events (pulling the handle, releasing the handle, and turning the key),
but what of a system with hundreds or thousands of times more?
Unfortunately, the number of possible interleavings grows {\em exponentially} with the number of events.

@@

\section{Contributions}

@@

The first half of the statement will serve as the overarching theme of this work:
that as impossible a problem as perfect formal verification may be,
one may still hope to achieve meaningful test results
using an algorithm that knows its own limits,
and compensates for them with heuristics informed by user's own human intuition.
The second half of the statement can be broken down into three parts,
coping with exponential explosion,
helping students,
and addressing modern concurrency models,
which correspond to the three major contributions of this thesis, as follows:}

\begin{enumerate}
	\item \revisionminor{{\bf Data-race preemption points and Iterative Deepening (\Cref{chap:quicksand}).}
		I will present Quicksand, a stateless model checking execution framework
		that manages multiple simultaneous Landslide instances
		to automatically cope with exponential explosion.
		Quicksand is powered by {\em Iterative Deepening},}
		a new algorithm for navigating the trade-off in how many preemption points to test at once.
		Iterative Deepening incorporates state space estimation \cite{estimation}
		to decide on-the-fly whether each state space is worth pursuing,
		and uses data race analysis \cite{tsan,fasttrack}
		to find new preemption point candidates based on a program's dynamic behaviour.
		\revision{This chapter will include a soundness proof,
		showing that testing only synchronization and data-race preemption points
		still constitutes a full formal verification of the test,
		and}
		a large evaluation, comparing its performance to three \revisionminor{prior-work} approaches
		across 600+ unique tests.
		I will show that \revisionminor{Quicksand}
		outperforms prior work in terms \revisionminor{of both} finding bugs quickly
		and verifying correctness when no bug exists.
	\item {\bf Educational use (\Cref{chap:education}).}
		For the past \revisionminor{seven} semesters,
		I have offered a fully-automated version of Landslide to students in 15-410,
		CMU's undergraduate Operating System Design and Implementation class,
		for use as a debugging aid during the thread library project \cite{thrlib}.
		%Recently
		I have also extended Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
		In the two most recent semesters, I collaborated with Operating Systems course staff at the University of Chicago,
		\revisionminor{which uses Pintos,
		and at The Pennsylvania State University,
		which has recently adopted CMU's thread library project,}
		to provide
		%debugging feedback to their students.
		\revisionminor{their students an opportunity to benefit from Landslide as well.}
		At all three universities I then collected statistics on the numbers and types of bugs found,
		and surveyed students to understand the human experience.
		This section will present the study's results
		\revisionminor{and prove that stateless model checking is suitable for use} in an educational setting.
	\item {\bf Hardware transactional memory (HTM) (\Cref{chap:tm}).}
		\revision{I will introduce a new concurrency model
		for stateless model checkers to emulate HTM's execution semantics
		in terms of existing concurrency primitives.
		This model is accompanied by a soundness proof
		which allows checkers to avoid simulating conflict abort rollbacks,
		reducing the state space substantially,
		while preserving the formal verification guarantee.
		I have implemented a new transactional testing mode in Landslide,
		with support for the many different abort reasons HTM introduces,
		optional weak atomicity semantics for simulating software TM instead,
		and a ``retry set'' reduction algorithm to identify and prune new types of equivalences.
		The evaluation includes several real-world HTM programs and benchmarks,
		and shows that Landslide can both find bugs and verify correctness
		using the new model with reasonable performance.}
\end{enumerate}

@@

I have tried to make this document accessible to readers of all programming experience levels,
although some of the research being theoretical and several layers of abstractions deep,
I cannot promise all easy reading.
\Cref{chap:background}, Background, provides what I hope are friendly concrete examples
to help the reader feel comfortable with each level of intuition that upcoming algorithms will build upon.
These should suffice for the experiments and overall contributions,
if not necessarily the details of each algorithm or soundness proof.
In particular, \Cref{chap:education}, Education,
may be approached with no knowledge of concurrency or model checking,
taking it merely as a study of a magic new debugging tool in the classroom setting.
The more ambitious reader may proceed to the Landslide chapter's algorithm walkthroughs~(\cref{sec:landslide-algs}),
which should equip them to understand every detail herein.
Readers who are here only to skim and skip around should at least be aware of the glossary~(\cref{sec:glossary})
to help clarify any terminology confusion.

@@

>>>>

My favourite way to explain concurrency to a non-technical audience
%(e.g., family members)
is by analogy with unlocking car doors.
In many %makes and
models of cars, the door handle and locking mechanism are intertwined
in such a way that the door cannot unlock while its handle is being pulled~\cite{car-doors}.
Most car users, then,
%are familiar with
have suffered the experience of
trying to open the passenger door
at exactly the same moment as the driver turns the key in the lock:
%not only does the door not open, but
it fails to unlock, and
they must beseech the driver to turn the key again!
If they had pulled the handle any later, it would have opened as normal;
any earlier, and it would have been as though not having the key at all,
but would at least have unlocked properly at the first turn of the key later.
One would reasonably expect to never have to turn the key more than once,
so this violation of expected behaviour is a {\em bug},
% more ambiguous version pointed out by garth
%This violation of expected behaviour -- that the key need not be turned twice -- is a {\em bug},
and its dependence on the two travelers' actions {\em interleaving}
%with the driver's actions
in a particular way
makes it a {\em concurrency bug}.
This example involves only three events (pulling the handle, releasing the handle, and turning the key),
but what of a system with hundreds or thousands of times more?
Unfortunately, the number of possible interleavings grows {\em exponentially} with the number of events.

@@

\section{Contributions}
\label{sec:thestatement}

@@

The first half of the statement will serve as the overarching theme of this work:
that as impossible a problem as perfect formal verification may be,
one may still hope to achieve meaningful test results
using an algorithm that knows its own limits,
and compensates for them with heuristics informed by the user's own human intuition.
The second half of the statement can be broken down into three parts:
coping with exponential explosion,
helping students,
and addressing modern concurrency models;
which correspond to the three major contributions of this thesis, as follows:}

\begin{enumerate}
	\item \revisionminor{{\bf Data-race preemption points and Iterative Deepening (\Cref{chap:quicksand}).}
		I will present Quicksand, a stateless model checking execution framework
		that manages multiple simultaneous Landslide instances
		to automatically cope with exponential explosion.
		Quicksand is powered by {\em Iterative Deepening},}
		a new algorithm for navigating the trade-off in how many preemption points to test at once.
		Iterative Deepening incorporates state space estimation \cite{estimation}
		to decide on-the-fly whether each state space is worth pursuing,
		and uses data race analysis \cite{tsan,fasttrack}
		to find new preemption point candidates based on a program's dynamic behaviour.
		\revision{This chapter will include a soundness proof,
		showing that testing only synchronization and data-race preemption points
		still constitutes a full formal verification of the test,
		and}
		a large evaluation, comparing its performance to three \revisionminor{prior-work} approaches
		across 600+ unique tests.
		I will show that \revisionminor{Quicksand}
		outperforms prior work in terms \revisionminor{of both} finding bugs quickly
		and verifying correctness when no bug exists.
	\item {\bf Educational use (\Cref{chap:education}).}
		For the past \revisionminor{seven} semesters,
		I have offered a fully-automated version of Landslide to students in 15-410,
		CMU's undergraduate Operating System Design and Implementation class,
		for use as a debugging aid during the thread library project \cite{thrlib}.
		%Recently
		I have also extended Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
		In the two most recent semesters, I collaborated with Operating Systems course staff at the University of Chicago,
		\revisionminor{which uses Pintos,
		and at The Pennsylvania State University,
		which recently adopted CMU's thread library project,}
		to provide
		%debugging feedback to their students.
		\revisionminor{their students an opportunity to benefit from Landslide as well.}
		At all three universities I then collected statistics on the numbers and types of bugs found,
		and surveyed students to understand the human experience.
		This section will present the study's results
		\revisionminor{and prove that stateless model checking is suitable for use} in an educational setting.
	\item {\bf Hardware transactional memory (HTM) (\Cref{chap:tm}).}
		\revision{I will introduce a new concurrency model
		for stateless model checkers to emulate HTM's execution semantics
		in terms of existing concurrency primitives.
		This model is accompanied by a soundness proof
		which allows checkers to avoid simulating conflict abort rollbacks,
		reducing the state space substantially,
		while preserving the formal verification guarantee.
		I have implemented a new transactional testing mode in Landslide,
		with support for the many different abort reasons HTM introduces,
		optional weak atomicity semantics for simulating software TM instead,
		and a ``retry set'' reduction algorithm to identify and prune new types of equivalences.
		The evaluation includes several real-world HTM programs and benchmarks,
		and shows that Landslide can both find bugs and verify correctness
		using the new model with reasonable performance.}
\end{enumerate}

@@

I have tried to make this document accessible to readers of all programming experience levels,
although some of the research being theoretical and several layers of abstractions deep,
I cannot promise easy reading to all.
\Cref{chap:background}, Background, provides what I hope are friendly concrete examples
to help the reader feel comfortable with each level of intuition that upcoming algorithms will build upon.
These should suffice for the experiments and overall contributions,
if not necessarily the details of each algorithm or soundness proof.
In particular, \Cref{chap:education}, Education,
may be approached with no knowledge of concurrency or model checking,
taking it merely as a study of a magic new debugging tool in the classroom setting.
The more ambitious reader may proceed to the Landslide chapter's algorithm walkthroughs~(\cref{sec:landslide-algs}),
which should equip them to understand every detail herein.
Readers who are here only to skim and skip around should at least be aware of the glossary~(\cref{sec:glossary})
to help clarify any terminology confusion.

@@

\subsubsection{Fonts}

The body of this thesis is set in Bitstream Charter,
the monospace font is Inconsolata,
the epigraph font is Alexa, %Std.
the Japanese font is Heisei Mincho, %Std W5.
and
the Arabic font is Amiri. %Regular.
%The SIGBOVIK font is Birbaslo.
